---
title: "Iteration"
output: html_notebook
---
```{r}
library(tidyverse) # includes tibbles
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

median(df$a) # [1] 0.3255279
median(df$b) # [1] -0.2812876
median(df$c) # [1] -0.2498801
median(df$d) # [1] -0.374399 why are these different??

output <- vector("double", ncol(df))  # 1. output
for (i in seq_along(df)) {            # 2. sequence
  output[[i]] <- median(df[[i]])      # 3. body
}

output # [1]  0.3255279 -0.2812876 -0.2498801 -0.3743990

y <- vector("double", 0)
seq_along(y) # integer(0)
1:length(y) # [1] 1 0
```
 Reduce duplication in your code by creating functions instead of copying-and-pasting!
 
Every for loop has three components:

The output: output <- vector("double", length(x)). Before you start the loop, you must always allocate sufficient space for the output. This is very important for efficiency: if you grow the for loop at each iteration using c() (for example), your for loop will be very slow.

A general way of creating an empty vector of given length is the vector() function. It has two arguments: the type of the vector (“logical”, “integer”, “double”, “character”, etc) and the length of the vector.

The sequence: i in seq_along(df). This determines what to loop over: each run of the for loop will assign i to a different value from seq_along(df). It’s useful to think of i as a pronoun, like “it”.

Write for loops to:

Compute the mean of every column in mtcars.
Determine the type of each column in nycflights13::flights.
Compute the number of unique values in each column of iris.
Generate 10 random normals for each of μ = −10, 0, 10, and 100

Think about the output, sequence, and body before you start writing the loop.

```{r}
output1 <- vector("double", ncol(mtcars))  # 1. output
for (i in seq_along(df)) {            # 2. sequence
  output[[i]] <- median(mtcars[[i]])      # 3. body
}
output # [1]  19.2   6.0 196.3 123.0
```

```{r}
#Eliminate the for loop in each of the following examples by taking advantage of an existing function that works with vectors:

out <- ""
for (x in letters) {
  out <- stringr::str_c(out, x)
}

x <- sample(100)
sd <- 0
for (i in seq_along(x)) {
  sd <- sd + (x[i] - mean(x)) ^ 2
}
sd <- sqrt(sd / (length(x) - 1))

x <- runif(100)
out <- vector("numeric", length(x))
out[1] <- x[1]
for (i in 2:length(x)) {
  out[i] <- out[i - 1] + x[i]
}
```
Combine your function writing and for loop skills:
Credit: https://github.com/cimentadaj/R4DS-Solutions/blob/master/ch17.Rmd 
Write a for loop that prints() the lyrics to the children’s song “Alice the camel”.
```{r}
song <-
  "Alice the camel has five humps.
Alice the camel has five humps.
Alice the camel has five humps.
So go, Alice, go.
"
split_song <- str_split(song, "\\.")[[1]]
for (i in split_song) {
  cat(i)
}
```


Convert the nursery rhyme “ten in the bed” to a function. Generalise it to any number of people in any sleeping structure.
```{r}
nurs <- function(x) {
  
  empty_list <- vector("list", x)
  empty_list[[1]] <- '\n There was 1 in the bed \n And the little one said, "Alone at last!"'
    
  for (i in seq_len(x)[-1]) {
    the_str <-
      paste0('\n There were ', i,' in the bed \n And the little one said, \n "Roll over! Roll over!" \n So they all rolled over and one fell out \n')
    
    empty_list[[i]] <- the_str
  }
  
  for (i in rev(empty_list)) cat(i, sep = "\n")
}
nurs(10)
```


Convert the song “99 bottles of beer on the wall” to a function. Generalise to any number of any vessel containing any liquid on any surface.
```{r}
beers <- function(x, drink, where) {
  
  for (i in c(rev(seq_len(x)), 0)) {
    if (i == 0) {
      cat(glue::glue("No more bottles of {drink} on the {where}, no more bottles of {drink}.\n Go to the store and buy some more, {x} bottles of {drink} on the {where}."))
    } else {
      cat(glue::glue("{i} {if (i == 1) 'bottle' else 'bottles'} of {drink} on the {where}, {i} {if (i == 1) 'bottle' else 'bottles'} of {drink}.\n Take one down and pass it around, {if (i == 1) 'no more' else i - 1} bottles of {drink} on the {where}. \n \n \n"))
    }
  }
}
beers(10, 'beer', 'wall')
beers(10, 'pop', 'floor')
```


It’s common to see for loops that don’t preallocate the output and instead increase the length of a vector at each step:
```{r}
output <- vector("integer", 0)
for (i in seq_along(x)) {
  output <- c(output, lengths(x[[i]]))
}
output # all 1s
```
How does this affect performance? Design and execute an experiment.
Growing the vector is slower than allocating.
```{r}
# use a for loop to modify an existing object
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}

df$a <- rescale01(df$a)
df$b <- rescale01(df$b)
df$c <- rescale01(df$c)
df$d <- rescale01(df$d)

# which can be written as
for (i in seq_along(df)) {
  df[[i]] <- rescale01(df[[i]])
}
```
A while loop is simpler than for loop because it only has two components, a condition and a body:

while (condition) {
  # body
}

A while loop is also more general than a for loop, because you can rewrite any for loop as a while loop, but you can’t rewrite every while loop as a for loop:

```{r}
for (i in seq_along(x)) {
  # body
}

# Equivalent to
i <- 1
while (i <= length(x)) {
  # body
  i <- i + 1 
```

Imagine you have a directory full of CSV files that you want to read in. You have their paths in a vector, files <- dir("data/", pattern = "\\.csv$", full.names = TRUE), and now want to read each one with read_csv(). Write the for loop that will load them into a single data frame.
```{r}
all_csv <- c("one.csv", "two.csv")
all_dfs <- vector("list", length(all_csv))
for (i in all_csv) {
  all_dfs[[i]] <- read_csv(all_csv[[i]])
}
bind_rows(all_dfs)
```

What happens if you use for (nm in names(x)) and x has no names? What if only some of the elements are named? What if the names are not unique?
```{r}
no_names <- 1:5
some_names <- c("one" = 1, 2, "three" = 3)
repeated_names <- c("one" = 1, "one" = 2, "three" = 3)
for (nm in names(no_names)) print(identity(nm)) # nothing happens!
for (nm in names(some_names)) print(identity(nm)) # the empty name get's filled with a ""
for (nm in names(repeated_names)) print(identity(nm)) # everything get's printed out
```

Write a function that prints the mean of each numeric column in a data frame, along with its name. For example, show_mean(iris) would print:

show_mean(iris)
# Sepal.Length: 5.84
# Sepal.Width:  3.06
# Petal.Length: 3.76
# Petal.Width:  1.20
(Extra challenge: what function did I use to make sure that the numbers lined up nicely, even though the variable names had different lengths?)
`cat` with `fill` set to `TRUE`
```{r}
show_means <- function(x) {
  
  the_class <- vector("logical", length(x))
  for (i in seq_along(x)) the_class[[i]] <- is.numeric(x[[i]])
  
  x <- x[the_class]
  
  for (i in seq_along(x)) {
    cat(paste0(names(x)[i], ": ", round(mean(x[[i]]), 2)), fill = TRUE)
  }
}
show_means(iris)
show_means(mtcars)
```

What does this code do? How does it work?

```{r}
trans <- list( 
  disp = function(x) x * 0.0163871,
  am = function(x) {
    factor(x, labels = c("auto", "manual"))
  }
)
for (var in names(trans)) {
  mtcars[[var]] <- trans[[var]](mtcars[[var]])
```
It converts `disp` and `am` by multiplying and then into a factor respectively. This is simply iterating over a list with functions, and applying in that same order to both variables.

Read the documentation for apply(). In the 2d case, what two for loops does it generalise?
```{r}
?apply
```

Adapt col_summary() so that it only applies to numeric columns You might want to start with an is_numeric() function that returns a logical vector that has a TRUE corresponding to each numeric column.
```{r}
col_summary <- function(x) {
  
  the_numeric <- vector("logical", length(x))
  for (i in seq_along(x)) the_numeric[[i]] <- is.numeric(x[[i]])
  
  x <- x[the_numeric]
  
  the_mean <- vector("numeric", length(x))
  for (i in seq_along(x)) the_mean[[i]] <- round(mean(x[[i]]), 2)
  
  the_mean
}
col_summary(iris)
```

There is one function for each type of output:

map() makes a list.
map_lgl() makes a logical vector.
map_int() makes an integer vector.
map_dbl() makes a double vector.
map_chr() makes a character vector.

```{r}
f <- c("runif", "rnorm", "rpois")
param <- list(
  list(min = -1, max = 1), 
  list(sd = 5), 
  list(lambda = 10)
)

#To handle this case, you can use invoke_map():
invoke_map(f, param, n = 5) %>% str()
# List of 3
#>  $ : num [1:5] 0.762 0.36 -0.714 0.531 0.254
#>  $ : num [1:5] 3.07 -3.09 1.1 5.64 9.07
#>  $ : int [1:5] 9 14 8 9 7
```

```{r}
dfs <- list(
  age = tibble(name = "John", age = 30),
  sex = tibble(name = c("John", "Mary"), sex = c("M", "F")),
  trt = tibble(name = "Mary", treatment = "A")
)

dfs %>% reduce(full_join)
```

Implement your own version of every() using a for loop. Compare it with purrr::every(). What does purrr’s version do that your version doesn’t?
```{r}
my_every <- function(.x, .f, ...) {
  the_test <- vector("logical", length(.x))
  for (i in seq_along(.x)) the_test[[i]] <- .f(.x[[i]], ...)
  all(the_test, na.rm = TRUE)
}
my_every(mtcars, is.numeric)
every(mtcars, is.numeric)
#`every` is probably much much faster and allows to supply one-sided formulas as functions
```

Create an enhanced col_summary() that applies a summary function to every numeric column in a data frame.
```{r}
col_sum <- function(x, f, ...) {
  x <- keep(x, is.numeric)
  map_dbl(x, f, ...)
}
col_sum(mtcars, mean, trim = 0.8)
col_sum(iris, median)
```

A possible base R equivalent of col_summary() is:

col_sum3 <- function(df, f) {
  is_num <- sapply(df, is.numeric)
  df_num <- df[, is_num]

  sapply(df_num, f)
}
But it has a number of bugs as illustrated with the following inputs:
```{r}
df <- tibble(
  x = 1:3, 
  y = 3:1,
  z = c("a", "b", "c")
)
# OK
col_sum3(df, mean)
# Has problems: don't always return numeric vector
col_sum3(df[1:2], mean)
col_sum3(df[1], mean)
col_sum3(df[0], mean)
```
What causes the bugs?
`sapply` returns different things based on input. `sapply(df[0], is.numeric)` returns a named list rather than a `tibble`, whereas `sapply(df[1], is.numeric)` returns the correct output.